A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-03-24 paper.tex
Some minor fixes for reviewer comments

    Reactive synthesis techniques based on constructing the winning region of
    the system have been shown to work well in many cases but suffer from state
    explosion in others.  A different approach, proposed recently, applies SAT
    solvers in a counterexample guided framework to solve the synthesis
    problem.  However, this method is limited to synthesising systems that
    execute for a bounded number of steps and is incomplete for synthesis with
    unbounded safety and reachability objectives.  We present an extension of
    this technique to unbounded synthesis.  Our method applies Craig
    interpolation to abstract game trees produced by counterexample-guided
    search in order to construct a monotonic sequence of may-losing regions.
    Experimental results based on SYNTCOMP 2015 competition benchmarks show
    this to be a promising alternative that solves some previously intractable
    instances.

@@

The reactive synthesis problem is 2EXPTIME-complete so
naive algorithms are infeasible on anything but simple systems.
% In this way the environment is equivalent to an existential search of
%the game and the controller is universal. Much of the complexity of reactive
%synthesis stems from managing the interactions of the alternating quantifiers
%on the state space of the game.
There are several techniques that aim to mitigate this complexity by
representing states and transitions of the system symbolically~\cite{Piterman_PS_06,bloem2014,morgenstern2013}. 
These techniques incrementally construct a symbolic representation of the set of 
discovered winning states of the game.  The downside of this approach is that 
keeping track of all discovered winning states can lead to a space explosion even
when using efficient symbolic representation such as BDDs or CNFs.

@@

An alternative approach, proposed by Narodytska et al.~\cite{narodytska2014},
is to eschew states and focus on \emph{runs} of the game.  The method works by
exploring a subset of the concrete runs of the game and proving that these runs can be
generalised into a winning strategy on behalf of one of the players. In contrast to 
other existing synthesis methods, it does not store,
in either symbolic or explicit form, the set of winning states.
Instead, it uses counterexample-guided backtracking search to identify a small subset
of runs that are sufficient to solve the game.

This method has been shown to outperform BDDs on certain classes of games; however it
suffers from an important limitation: it is only able to solve games with a bounded number
of rounds.  In case of safety games, this means proving that the controller 
can keep the game within the safe region for a bounded number of steps.  
This is insufficient in most practical situations that require unbounded realizability.

%Previous work has applied this idea to
%realizability of bounded games \cite{narodytska2014} by forming abstractions of
%the game and refining in a counterexample-guided framework.

In this paper, we extend the method by Narodytska et al.~\cite{narodytska2014} to unbounded 
safety games.  To this end we enhance the method with computational learning: every time
the search algorithm discovers a new counterexample, the learning procedure analyzes the 
counterexample, extracting a subset of states winning for one of the players from it.  
The learning procedure ensures that reaching a fixed point in these sets is sufficient to establish 
unbounded realizability.
Our method can be seen as a hybrid between the counterexample-guided algorithm
by Narodytska et al. and methods based on losing set computation: we use the
former to guide the search, while we rely on the latter to ensure convergence.

@@

A \emph{controller strategy} $\pi^c : (X, U) \to C$ is a mapping of states and
uncontrollable inputs to controllable actions. A controller strategy is winning
in a bounded game of maximum bound $\kappa$ if all runs $(x_0, u_0, \pi^c(x_0,
u_0)), (x_1, u_1, \pi^c(x_1, u_1)) \dots$ $(x_n, u_n, \pi^c(x_n, u_n))$ are
winning.  Bounded \emph{realizability} is the problem of determining the
existence of such a strategy for a bounded game.

An \emph{environment strategy} $\pi^e : X \to U$ is a mapping of states to
uncontrollable actions. A bounded run is winning for the environment if $x_0 =
I \land \exists i \in \{1..n\} (E(x_i))$ and an environment strategy is winning
for a bounded game if all runs $(x_0, \pi^e(x_1), c_1), (x_1, \pi^e(x_1), c_1)
\dots (x_n, \pi^e(x_n), c_n)$ are winning for the environment.  Safety games
are zero sum, therefore the existence of a winning controller strategy implies
the nonexistence of a winning environment strategy and vice versa.

\subsection{Counterexample-guided bounded synthesis}

@@

An abstraction of the game restricts actions available to one of the players.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees}.  Figure~\ref{fig:agt} shows an
example abstract game tree for the controller (abstract game trees for the
environment are similar).  In the abstract game, the controller is required to
pick actions from the tree, starting from the root node.  After reaching a
leaf, it continues playing unrestricted.  The tree in Figure~\ref{fig:agt}
restricts the initial controller action to the set $\{a, d\}$.  After choosing
action $a$, the controller is required to play action $b$ in the next round.
It then reaches a leaf of the tree and continues playing unrestricted.  The
root of the tree is annotated by the initial state $s$ of the abstract game and
the bound $k$ on the number of rounds.  We denote $\textsc{nodes}(T)$ the set
of all nodes of a tree $T$, $\textsc{leaves}(T)$ the subset of leaf nodes.  For
edge $e$, $\textsc{action}(e)$ is the action that labels the edge, and for node
$n$, $\textsc{height}(n)$ is the maximal distance from n to a leaf. For node
$n$ of the tree, $\textsc{succ}(n)$ is the set of pairs $\langle e, n' \rangle$
where $n'$ is a child node of $n$ and $e$ is the edge connecting $n$ and $n'$.

@@

The bounded synthesis algorithm solves abstract game trees by invoking a SAT
solver to search for candidate partial strategies. Candidate strategies are
checked and by a call to the opponent solver, which also refines the abstract
game tree. The refinements are then solved recursively.  The full procedure is
illustrated in Algorithm~\ref{alg:bounded}.

\begin{algorithm}[t]
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$p, s, k, T$}
        \State $cand \gets $ \Call{findCandidate}{$p, s, k, T$} \Comment{Look for a candidate}
        \IIf{$k = 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, s, k, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{append}{$T', l, u$} \Comment{Refine $T'$ with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, s, k, T'$} \Comment{Solve refined game tree}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, s, k, T$}
        \State $\hat{T} \gets $ \Call{extend}{$T$} \Comment{Extend the tree with unfixed actions}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$\hat{T}$}}{\Call{\textoverline{treeFormula}}{$\hat{T}$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_{\hat{T}}) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \If{\texttt{unbounded}} \Comment{Active only in the unbounded solver}
                    %\State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                    \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$s, \hat{T}$}}{\Call{\textoverline{learn}}{$s, \hat{T}$}} \EndIfElse
                \EndIf
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{nodes}{$T$} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b2}
        \Function{verify}{$p, s, k, T, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$s, k, cand, l$} \Comment{Get bound and state at leaf}
                \State $u \gets $ \Call{solveAbstract}{\Call{opponent}{$p$}, $k'$, $s'$, $\emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

The tree is first extended to the maximum bound with edges that are labeled
with arbitrary opponent actions (Algorithm~\ref{alg:bounded}, line 14) as shown
in Figure~\ref{fig:extended}. Then copies of state and action variables are
introduced for each node in tree and opponent action variables for each edge
$e$ set to $\textsc{action}(e)$. The state and action variables of node $n$ are
connected to successor nodes $\textsc{succ}(n)$ by an encoding of the
transition relation and constrained to the winning condition of the player.

@@

\begin{algorithm}
    \caption{Tree formulas for Controller and Environment respectively}
    \label{alg:treeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{$T$}
        \If{$\Call{height}{T} = 0$}
        \State \Return{ $\lnot \Call{E}{X_{T}}$ }
        \Else
        \State \Return{$\lnot \Call{E}{X_{T}} \land$ \\
            $$\bigwedge_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_e, C_n, X_n} \land U_e = \Call{action}{e} \land \Call{treeFormula}{n})$$
        }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{$T$}
        \If{$\Call{height}{T} = 0$}
        \State \Return{\Call{E}{$X_{T}$}}
        \Else
        \State \Return{ $\Call{E}{X_{T}} \lor$ \\
        $$\bigvee_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_n, C_e, X_n} \land C_e = \Call{action}{e} \land \Call{\textoverline{treeFormula}}{n})$$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

We extend the bounded synthesis algorithm to learn states losing for one of the players from
failed attempts to find candidate strategies.  The learning procedure kicks in whenever
\textsc{findCandidate} cannot find a candidate strategy for an abstract game tree with initial
state $s$.  At this point we have established that $s$ is a losing state.  We can learn additional losing
states from the tree via interpolation.  This is achieved in lines~18--20 in Algorithm~\ref{alg:bounded},
enabled in the unbounded version of the algorithm, which invoke \textsc{learn} or
\textsc{\textoverline{learn}} to learn controller or environment losing states respectively
(Algorithm~\ref{alg:learn}).

\begin{figure}[t]
    \centering
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left] {$e$}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right] {$f$}
                    }
                    edge from parent [fixed] {node [right] {$d$}}
                }
                node [above=4pt] {$\langle s, 2 \rangle$};
            \node [right=1pt of n] {$n$};
        \end{tikzpicture}
        \caption{A losing AGT $T$}
        \label{fig:interpolatetree}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left] {$e$}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right] {$f$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, 2 \rangle$};
            \node [right=1pt of n] {$n$};
        \end{tikzpicture}
        \caption{Tree slice $T_1$}
        \label{fig:treef1}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] {}
                child {node [circle,draw,right=5pt] {}
                    edge from parent [fixed] node [left] {$e$}
                }
                child {node [circle,draw,left=5pt] {}
                    edge from parent [fixed] node [right] {$f$}
                }
                node [right=5pt] {$n$};
        \end{tikzpicture}
        \caption{Tree slice $T_2$}
        \label{fig:treef2}
    \end{subfigure}
    \caption{Splitting of an abstract game tree by the learning procedure.}
\end{figure}

\begin{algorithm}[t]
    \begin{algorithmic}[1]
        \Require $s(X_T) \land $ \Call{treeFormula}{$T$} $\equiv \bot$
        \Require \emph{Must-invariant} holds
        \Ensure \emph{Must-invariant} holds
        \Ensure $s(X_T) \land B^M$
        \Function{learn}{$s, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min height 
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$s(X_T) \land $ \Call{treeFormula}{$T_1$}, \Call{treeFormula}{$T_2$}}
            \State $B^M \gets B^M \lor \mathcal{I}$
            \State \Call{learn}{$s, T_1$}
        \EndFunction
        \algstore{learn}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{learn}
        \Require $s(X_T) \land $ \Call{\textoverline{treeFormula}}{$T$} $\equiv \bot$
        \Require \emph{May-invariant} holds
        \Ensure \emph{May-invariant} holds
        \Ensure $s(X_T) \land B^m[$\Call{height}{$T$}$] \equiv \bot$
        \Function{\textoverline{learn}}{$s, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min height 
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$s(X_T) \land $ \Call{\textoverline{treeFormula}}{$T_1$}, \Call{\textoverline{treeFormula}}{$T_2$}}
            \For{$i = 1$ to \Call{height}{$n$}}
                \State $B^m[i] \gets B^m[i] \setminus \mathcal{I}$
            \EndFor
            \State \Call{\textoverline{learn}}{$s, T_1$}
        \EndFunction
    \end{algorithmic}
    \caption{Learning algorithms}
    \label{alg:learn}
\end{algorithm}

@@

We have discovered a set $\mathcal{I}$ of states losing for the controller.  
We record these newly discovered states in a set $B^M$ of
``bad'' states that \emph{must} be losing for the controller (Algorithm~\ref{alg:learn}, line~6). This set is initialised to the error
set $E$ and grows as new losing states are
discovered.  The \textsc{treeFormula} function is modified for the unbounded
solver (Algorithm~\ref{alg:unboundedTreeFormula}) to replace $E$ with
$B^M$. This enables further interpolants to be constructed by the learning procedure 
recursively splitting more nodes from $T_1$ (Algorithm~\ref{alg:learn}, line~7) since the states that lost to $T_2$ are now
blocked by $B^M$.

@@

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{$T$}
        \If{$\Call{height}{T} = 0$}
        \State \Return{\Call{E}{$X_T$}}
        \Else
        \State \Return{ \Call{$B^m[\Call{height}{T}]$}{$X_{T}$} $\land$ \\
            $$\Call{E}{X_T} \lor \bigvee_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_n, C_e, X_n} \land C_e = \Call{action}{e} \land \Call{\textoverline{treeFormula}}{n})$$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

>>>>

    Reactive synthesis techniques based on constructing the winning region of
    the system have been shown to work well in many cases but suffer from state
    explosion in others.  A different approach, proposed recently, applies SAT
    solvers in a counterexample guided framework to solve the synthesis
    problem.  However, this method is limited to synthesising systems that
    execute for a bounded number of steps and is incomplete for synthesis with
    unbounded safety and reachability objectives.  We present an extension of
    this technique to unbounded synthesis.  Our method applies Craig
    interpolation to abstract game trees produced by counterexample guided
    search in order to construct a monotonic sequence of may-losing regions.
    Experimental results based on SYNTCOMP 2015 competition benchmarks show
    this to be a promising alternative that solves some previously intractable
    instances.

@@

The reactive synthesis problem is EXPTIME-complete so
naive algorithms are infeasible on anything but simple systems.
% In this way the environment is equivalent to an existential search of
%the game and the controller is universal. Much of the complexity of reactive
%synthesis stems from managing the interactions of the alternating quantifiers
%on the state space of the game.
There are several techniques that aim to mitigate this complexity by
representing states and transitions of the system symbolically~\cite{Piterman_PS_06,bloem2014,morgenstern2013}. 
These techniques incrementally construct a symbolic representation of the set of 
discovered winning states of the game.  The downside of this approach is that 
keeping track of all discovered winning states can lead to a space explosion even
when using efficient symbolic representation such as BDDs or CNFs.

@@

An alternative approach, proposed by Narodytska et al.~\cite{narodytska2014},
is to eschew states and focus on \emph{runs} of the game.  The method works by
exploring a subset of the concrete runs of the game and proving that these runs can be
generalised into a winning strategy on behalf of one of the players. In contrast to 
other existing synthesis methods, it does not store,
in either symbolic or explicit form, the set of winning states.
Instead, it uses counterexample guided backtracking search to identify a small subset
of runs that are sufficient to solve the game.

This method has been shown to outperform BDDs on certain classes of games; however it
suffers from an important limitation: it is only able to solve games with a bounded number
of rounds.  In case of safety games, this means proving that the controller 
can keep the game within the safe region for a bounded number of steps.  
This is insufficient in most practical situations that require unbounded realisability.

%Previous work has applied this idea to
%realisability of bounded games \cite{narodytska2014} by forming abstractions of
%the game and refining in a counterexample-guided framework.

In this paper, we extend the method by Narodytska et al.~\cite{narodytska2014} to unbounded 
safety games.  To this end we enhance the method with computational learning: every time
the search algorithm discovers a new counterexample, the learning procedure analyzes the 
counterexample, extracting a subset of states winning for one of the players from it.  
The learning procedure ensures that reaching a fixed point in these sets is sufficient to establish 
unbounded realisability.
Our method can be seen as a hybrid between the counterexample guided algorithm
by Narodytska et al. and methods based on losing set computation: we use the
former to guide the search, while we rely on the latter to ensure convergence.

@@

A \emph{controller strategy} $\pi^c : (2^X, 2^U) \to 2^C$ is a mapping of states and
uncontrollable inputs to controllable actions. A controller strategy is winning
in a bounded game of maximum bound $\kappa$ if all runs $(x_0, u_0, \pi^c(x_0,
u_0)), (x_1, u_1, \pi^c(x_1, u_1)) \dots$ $(x_n, u_n, \pi^c(x_n, u_n))$ are
winning.  Bounded \emph{realisability} is the problem of determining the
existence of such a strategy for a bounded game.

An \emph{environment strategy} $\pi^e : 2^X \to 2^U$ is a mapping of states to
uncontrollable actions. A bounded run is winning for the environment if $x_0 =
I \land \exists i \in \{1..n\} (E(x_i))$ and an environment strategy is winning
for a bounded game if all runs $(x_0, \pi^e(x_0), c_0), (x_1, \pi^e(x_1), c_1)
\dots (x_n, \pi^e(x_n), c_n)$ are winning for the environment.  Safety games
are zero sum, therefore the existence of a winning controller strategy implies
the nonexistence of a winning environment strategy and vice versa.

\subsection{Counterexample guided bounded synthesis}

@@

An abstraction of the game restricts actions available to one of the players.
Specifically, we consider abstractions represented as trees of actions,
referred to as \emph{abstract game trees} (AGTs).  Figure~\ref{fig:agt} shows an
example abstract game tree for the controller (abstract game trees for the
environment are similar).  In the abstract game, the controller is required to
pick actions from the tree, starting from the root node.  After reaching a
leaf, it continues playing unrestricted.  The tree in Figure~\ref{fig:agt}
restricts the initial controller action to the set $\{a, d\}$.  After choosing
action $a$, the controller is required to play action $b$ in the next round.
It then reaches a leaf of the tree and continues playing unrestricted.  The
root of the tree is annotated by the initial state $s$ of the abstract game and
the bound $k$ on the number of rounds.  We denote $\textsc{nodes}(T)$ the set
of all nodes of a tree $T$, $\textsc{leaves}(T)$ the subset of leaf nodes.  For
edge $e$, $\textsc{action}(e)$ is the action that labels the edge, and for node
$n$, $\textsc{height}(n)$ is the maximal distance from n to a leaf. $\textsc{height}(T)$ is the height 
of the root node of the tree.  For node
$n$ of the tree, $\textsc{succ}(n)$ is the set of pairs $\langle e, n' \rangle$
where $n'$ is a child node of $n$ and $e$ is the edge connecting $n$ and $n'$.

@@

The bounded synthesis algorithm solves abstract game trees by invoking a SAT
solver to search for candidate partial strategies. Candidate strategies are
checked by a call to the opponent solver, which also refines the abstract
game tree. The refinements are then solved recursively.  The full procedure is
illustrated in Algorithm~\ref{alg:bounded}.

\begin{algorithm}[t]
    \begin{algorithmic}[1]
        \Function{solveAbstract}{$p, s, k, T$}
        \State $cand \gets $ \Call{findCandidate}{$p, s, k, T$} \Comment{Look for a candidate}
        \IIf{$k = 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \texttt{NULL}$} \Return $\texttt{NULL}$ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, s, k, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{append}{$T', l, u$} \Comment{Refine $T'$ with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, s, k, T'$} \Comment{Solve refined game tree}
        \EndLoop
        \EndFunction
        \algstore{b1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, s, k, T$}
        \State $\hat{T} \gets $ \Call{extend}{$T$} \Comment{Extend the tree with unfixed actions}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$\hat{T}$}}{\Call{\textoverline{treeFormula}}{$\hat{T}$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_{\hat{T}}) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \If{\texttt{unbounded}} \Comment{Active only in the unbounded solver}
                    %\State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                    \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$s, \hat{T}$}}{\Call{\textoverline{learn}}{$s, \hat{T}$}} \EndIfElse
                \EndIf
                \State \Return $\texttt{NULL}$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{nodes}{$T$} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{b2}
        \Function{verify}{$p, s, k, T, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$s, k, cand, l$} \Comment{Get bound and state at leaf}
                \State $u \gets $ \Call{solveAbstract}{\Call{opponent}{$p$}, $s'$, $k'$, $\emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \texttt{NULL}$} \Return $\langle \True, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \False, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

@@

The tree is first extended to the maximum bound with edges that are labeled
with arbitrary opponent actions (Algorithm~\ref{alg:bounded}, line 14) as shown
in Figure~\ref{fig:extended}.  For each node in the tree, new SAT variables are
introduced corresponding to the state ($X_T$) and action ($U_T$ or $C_T$)
variables of that node. Additional variables for the opponent actions in the
edges of $T$ are introduced ($U_e$ or $C_e$) and set to $\textsc{action}(e)$.
The state and action variables of node $n$ are connected to successor nodes
$\textsc{succ}(n)$ by an encoding of the transition relation and constrained to
the winning condition of the player.

%%%Then copies of state and action variables are introduced for each node in the
%%%tree and opponent action variables for each edge $e$ are set to
%%%$\textsc{action}(e)$.

@@

\begin{algorithm}
    \caption{Tree formulas for Controller and Environment respectively}
    \label{alg:treeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{$T$}
        \If{$\Call{height}{T} = 0$}
        \State \Return{ $\lnot \Call{E}{X_{T}}$ }
        \Else
        \State \Return{$\lnot \Call{E}{X_{T}} \land$ \\
            $$\bigwedge_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_e, C_T, X_n} \land U_e = \Call{action}{e} \land \Call{treeFormula}{n})$$
        }
        \EndIf
        \EndFunction
        \algstore{tf1}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{$T$}
        \If{$\Call{height}{T} = 0$}
        \State \Return{\Call{E}{$X_{T}$}}
        \Else
        \State \Return{ $\Call{E}{X_{T}} \lor$ \\
        $$\bigvee_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_T, C_e, X_n} \land C_e = \Call{action}{e} \land \Call{\textoverline{treeFormula}}{n})$$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

We extend the bounded synthesis algorithm to learn states losing for one of the
players from failed attempts to find candidate strategies.  The learning
procedure kicks in whenever \textsc{findCandidate} cannot find a candidate
strategy for an abstract game tree $T$ with initial state $s$ and bound $k$.
For the system, this means the environment partial strategy represented by $T$
will always reach $E$ from $s$ no matter the assignments to $C$ variables (the
bound is irrelevant).  For the environment, the system can avoid $E$ for $k$
rounds no matter the assignments to $U$ variables. Hence, we have established
that $s$ is a losing state either always (for the system) or for bound $k$ (for
the environment. We can learn additional losing states from the tree via
interpolation.  This is achieved in lines~18--20 in
Algorithm~\ref{alg:bounded}, enabled in the unbounded version of the algorithm,
which invoke \textsc{learn} or \textsc{\textoverline{learn}} to learn
controller or environment losing states respectively
(Algorithm~\ref{alg:learn}).

\begin{figure}[t]
    \centering
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left] {$e$}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right] {$f$}
                    }
                    edge from parent [fixed] {node [right] {$d$}}
                }
                node [above=4pt] {$\langle s, 2 \rangle$};
            \node [right=1pt of n] {$n$};
        \end{tikzpicture}
        \caption{A losing AGT $T$}
        \label{fig:interpolatetree}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] (n) {}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, 2 \rangle$};
            \node [right=1pt of n] {$n$};
        \end{tikzpicture}
        \caption{Tree slice $T_1$}
        \label{fig:treef1}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] {}
                child {node [circle,draw,right=5pt] {}
                    edge from parent [fixed] node [left] {$e$}
                }
                child {node [circle,draw,left=5pt] {}
                    edge from parent [fixed] node [right] {$f$}
                }
                node [right=5pt] {$n$};
        \end{tikzpicture}
        \caption{Tree slice $T_2$}
        \label{fig:treef2}
    \end{subfigure}
    \caption{Splitting of an abstract game tree by the learning procedure.}
\end{figure}

\begin{algorithm}[t]
    \begin{algorithmic}[1]
        \Require $s(X_T) \land $ \Call{treeFormula}{$T$} $\equiv \bot$
        \Require \emph{Must-invariant} holds
        \Ensure \emph{Must-invariant} holds
        \Ensure $s(X_T) \land B^M \not\equiv \bot$
        \Comment $s$ will be added to $B^M$
        \Function{learn}{$s, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min height 
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$s(X_T) \land $ \Call{treeFormula}{$T_1$}, \Call{treeFormula}{$T_2$}}
            \State $B^M \gets B^M \lor \mathcal{I}$
            \State \Call{learn}{$s, T_1$}
        \EndFunction
        \algstore{learn}
    \end{algorithmic}

    \begin{algorithmic}[1]
        \algrestore{learn}
        \Require $s(X_T) \land $ \Call{\textoverline{treeFormula}}{$T$} $\equiv \bot$
        \Require \emph{May-invariant} holds
        \Ensure \emph{May-invariant} holds
        \Ensure $s(X_T) \land B^m[$\Call{height}{$T$}$] \equiv \bot$
        \Comment{$s$ will be removed from $B^m[$\Call{height}{$T$}$]$}
        \Function{\textoverline{learn}}{$s, T$}
            \IIf{$T = \emptyset$}
                \Return
            \EndIIf
            \State $n \gets $ non-leaf node with min height 
            \State $\langle T_1, T_2 \rangle \gets $ \Call{gtSplit}{$T, n$}
            \State $\mathcal{I} \gets $ \Call{interpolate}{$s(X_T) \land $ \Call{\textoverline{treeFormula}}{$T_1$}, \Call{\textoverline{treeFormula}}{$T_2$}}
            \For{$i = 1$ to \Call{height}{$n$}}
                \State $B^m[i] \gets B^m[i] \setminus \mathcal{I}$
            \EndFor
            \State \Call{\textoverline{learn}}{$s, T_1$}
        \EndFunction
    \end{algorithmic}
    \caption{Learning algorithms}
    \label{alg:learn}
\end{algorithm}

@@

We have discovered a set $\mathcal{I}$ of states losing for the controller. We
record these newly discovered states in a set $B^M$ of ``bad'' states that
\emph{must} be losing for the controller (Algorithm~\ref{alg:learn}, line~6).
This set is initialised to the error set $E$ and grows as new losing states are
discovered.  The \textsc{treeFormula} function is modified for the unbounded
solver (Algorithm~\ref{alg:unboundedTreeFormula}) to replace $E$ with $B^M$.
This enables further interpolants to be constructed by the learning procedure
recursively splitting more nodes from $T_1$ (Algorithm~\ref{alg:learn}, line~7)
since the states that are losing to $T_2$ are now blocked by $B^M$.

@@

    \begin{algorithmic}[1]
        \algrestore{tf1}
        \Function{\textoverline{treeFormula}}{$T$}
        \If{$\Call{height}{T} = 0$}
        \State \Return{\Call{E}{$X_T$}}
        \Else
        \State \Return{ \Call{$B^m[\Call{height}{T}]$}{$X_{T}$} $\land$ \\
            $$\bigg( \Call{E}{X_T} \lor \bigvee_{\langle e, n \rangle \in \Call{succ}{T}}(\Call{$\delta$}{X_T, U_n, C_e, X_n} \land C_e = \Call{action}{e} \land \Call{\textoverline{treeFormula}}{n})\bigg)$$ }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

Learning of states losing for the environment is similar (\textsc{\textoverline{learn}} in Algorithm~\ref{alg:learn}). 
The main difference is that environment-losing states are only losing for a particular
bound: given that there does not exist an environment strategy that
forces the game into an error state in $k$ rounds or less; there may still exist a longer
environment-winning strategy.  We therefore record learned environment-losing states
along with associated bounds.  To this end, we maintain a conceptually infinite array of
sets $B^m[k]$ that are may-losing for the controller, indexed by bound $k$.  $B^m[k]$ 
are initialised to $\top$ for all $k$.  Whenever an environment-losing set $\mathcal{I}$
is discovered for a node $n$ with bound $\textsc{height}(n)$ in line~13 of Algorithm~\ref{alg:learn}, this set is subtracted from 
$B^m[i]$, for all $i$ less than or equal to the bound (lines~14--16).
